Epistemic Semantics in Guarded String Models

### 2020-09-27 tex/sdraft.tex
Edits in sdraft, nearly there!

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}
%% \newcommand{\ehc}[1]{}
%% \newcommand{\mr}[1]{}

@@

\usepackage{amssymb,amsmath,accents}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{hang}
\renewcommand{\UrlFont}{\ttfamily\small}
\usepackage{microtype}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
xs
{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except (depending on the Epik model) as
  an entailment.}

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $RS$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition (.o.), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.

@@

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
fusion product is lifted to lists instead of sets, negation is
implemented using list difference ($\setminus$), and the modal
operator lifts the alternative relation over lists of
strings\footnote{Figure~\ref{fig:list-interp} depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

Sentences are parsed with a chart parser for categorial grammar.  The
semantics for complex phrases are obtained by application of semantic
translations, accompanied by beta reductions that eliminate all
lambdas logical forms for clauses.  In consequence, the semantic
term translating a sentence is an Epik term.  Such a term designates a
set of possible words (guarded strings).  By way of example,
\exi{three}{a} is an English sentence with conjunction and several
levels of clausal embedding.  Using the grammar and parser, the
sentence is mapped to the term in \exi{three}{b}. \exi{three}{c}
shows a simplified logical from constructed form \exi{three}{b} using
logical equivalences. The origial term is compiled in an
implementation of the finite state calculus to a finite state machine
with ====== states and ====== arcs, which accepts a countably infinite set
of worlds. \ehc{Is this right anymore? the formula has changed
  drastically. MR: No, we need to write the Epik program and compile it
into Fst. The sentences might not be equivalent any more.}. In this way the methodology ``directly'' represents
the set of worlds denoted by \exi{three}{a}.

@@

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by the grammar.
Logical relations between propositions are checked in the finite state calculus by
checking set-theoretic relations between sets of worlds (as represented by finite
state machines in an interpreter for the finite state calculus).  In this case, the
propositions \exi{three}{b} and \exi{three}{e} are equivalent, in the sense that they
denote the same set of worlds.  This equivalence is attested computationally in the
possible worlds model.  Entailment can be tested with a subset check on sets of
worlds, and logical compatibility with a check for non-empty intersection.

@@

The model framework is a constructive branching-time framework with metaphysical
modality and epistemic modality, which should be applicable in linguistic semantic
research on combinations of tense, metaphysical modality, and epistemic
complementation \citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}.  Connections with research on temporal constitution of
events in a related formal setting remain to be explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.

The development here is concerned with defining concrete possible
worlds models, and applying them in natural language semantics.  
Issues for further investigation are mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms, coalgebra, and decidability.

>>>>

\newcommand{\ehc}[1]{\textit{\textcolor{red}{[ehc]: #1}}}
\newcommand{\mr}[1]{\textit{\textcolor{cyan}{[mr]: #1}}}
% \newcommand{\ehc}[1]{}
% \newcommand{\mr}[1]{}

@@

\usepackage{amssymb,amsmath,accents}
\usepackage{stmaryrd}
\SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}
\usepackage{array}
\usepackage{hang}
\renewcommand{\UrlFont}{\ttfamily\small}
\usepackage{microtype}

@@

\begin{example}   \label{interpretation1}
\hspace{-1.5em}\begin{tabular}[t]{ll}
${\bullet}^*t$ & It's tails. \\
$\bullet^*h$ & It's heads. \\  
  $ \bullet^* h \wedge \Box_a \bullet^* h$  & Amy knows it's heads. \\
\end{tabular}
\vspace{1mm}
\hspace{-1.5em}\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob believes Amy knows whether it's tails.  \\
\end{tabular}
\end{example}
xs
{\em Standard Epistemic Modalities}
Using our existential modal primitive $\Diamond_a$, and the dual
encodings of $\Box_a$ and $\wedge$, we can encode the standard modal
operators expressing knowledge $(\K_a)$ and belief $(\B_a)$
as in \ex{encodings}.\footnote{Deeper analyisis of the lexical semantics
  of {\em know} requires adding modeling of presupposition
  \citep{collard2018finite}. The grammar fragment in Section 6 does not model
  the presupposition of {\em know}, except as
  an entailment.}

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms. 
In Fst, we use the closure of the concatenation
product operation on relations to lift a relation on decorated events for an agent to
the corresponding relation on worlds.  The concatenation product $RS$ of two
relations $R$ and $S$ is the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where
$x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under discussion here is
  the closure of the former rather than the latter.}  The epistemic alternative
relation on worlds for an agent is then defined as the KAT relational concatenation
closure {\tt RelKst} of the decorated-event alternative relation for the agent.

@@

We lift the remaining operators (except Kleene star) to their list
equivalents: union becomes list append (written $\texttt{+\!+}$);
fusion product is lifted to lists instead of sets, negation is
implemented using list difference ($\setminus$), and the modal
operator lifts the alternative relation over lists of
strings\footnote{Figure~7 depicts this using the
  list comprehension notation, which is analogous to set builder
  notation, except that it is written using square brackets. Element
  order is evoked by the keyword \texttt{for}, rather than using the
  unordered $\forall$.}. The only caveat to these direct
interpretations is that we restrict the operators to have size $\leq
n$, denoted as $l|_n$ for a list of guarded strings $l$.

@@

Sentences are parsed with a chart parser for categorial grammar.  The semantics for
complex phrases are obtained by application of semantic translations, accompanied by
beta reductions that eliminate all lambdas logical forms for clauses.  In
consequence, the semantic term translating a sentence is an Epik term.  Such a term
designates a set of possible words (guarded strings).  By way of example,
\exi{three}{a} is an English sentence with conjunction and several levels of clausal
embedding.  Using the grammar and parser, the sentence is mapped to the term in
\exi{three}{b}. \exi{three}{c} shows a simplified logical from constructed form
\exi{three}{b} using logical equivalences. Either term is compiled in an
implementation of the finite state calculus to a finite state machine with 10
states and 23 edges, which accepts a countably infinite set of worlds.\footnote{
Machine sizes need not be small, especially as the cardinality of $\sf{B}$
increases.  A certain Epik model with fourteen primitive tests has the set of worlds
representented by a finite state machine with 184794 states and 257881
edges.}
In this way the methodology ``directly'' represents the set of worlds
denoted by \exi{three}{a}.

@@

\end{examples}
Sentence \exi{three}{d} is assigned the logical form \exi{three}{e} by the grammar.
Logical relations between propositions are checked in the finite state calculus by
checking set-theoretic relations between sets of worlds.
For instance entailment $p \rightarrow q$ is decided by checking in an interpreter
for the finite sate calculus whether $p - q$ is non-empty.
In the model defined by Figure 2, the
propositions \exi{three}{b} and \exi{three}{e} are
independent. They are equivalent in a version 
without secret flipping.

@@

The model framework is a constructive branching-time framework with metaphysical
modality and epistemic modality, which will be applicable in linguistic semantic
research on combinations of tense, metaphysical modality, and epistemic
complementation \citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}.  Connections with research on temporal constitution of
events in a related formal setting remain to be explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.

The development here is concerned with defining concrete computable possible
worlds models, and applying them in natural language semantics.  
Issues for further investigation are mathematical characterizations
of epistemic KATs, e.g. sound and complete axioms, coalgebra, and decidability.
